# Customise
 ----------------

## {#sec:customAP}

Customising the ageprior is generally not necessary, as a sample-specific ageprior will be estimated after parentage assignment based on the scaffold-pedigree. 

Parentage assignment is rarely improved by a non-default ageprior, unless the number of SNPs is so low or their quality so poor that it is difficult to genetically distinguish between parent-offspring and full sibling pairs. 

The ageprior used for full pedigree reconstruction may require tweaking if for example the assigned parents reflect only a portion of the biologically possible parental ages, or if some assigned parents are wrong and have a biologically impossible age.  



## Via `args.AP`
During pedigree reconstruction, you can pass arguments to `MakeAgePrior()` via  `sequoia()`'s argument `args.AP`. Use this to explicitly specify [`Discrete`](#sec:Discrete), [`Smooth`](#sec:Smooth) or [`Flatten`](#sec:Flatten), or to increase [`MaxAgeParent`](#sec:MaxAgePO) or `MinAgeParent`.  


## Different input pedigree
If you have a reasonable quality pedigree that contains many more individuals than have been SNP-genotyped, for example based on field observations and/or microsatellite data, an ageprior estimated from that old pedigree may be more informative than the one estimated from a limited number of SNP-genotyped parent-offspring pairs. A pedigree cannot be passed via `args.AP`, but this can be done as follows instead:

```{r APold, eval=FALSE}
APfromOld <- MakeAgePrior(Pedigree = MyOldPedigree,
                          LifeHistData = LH,
                          Smooth = TRUE)
SeqOUT <- sequoia(GenoM = Geno,
                  LifeHistData = LH,
                  SeqList = list(AgePriors = APfromOld))
```

The ageprior in `SeqList` will be used for both parentage assignment and full pedigree reconstruction. 


## Other tweaks
The easiest way to create a fully custom ageprior is to generate an approximate ageprior with `MakeAgePrior`, and tweak it by hand or using your own function. For example, imagine a species where females breed only every other year, but may have twins:
```{r skip, fig.cap="", out.width = "70%"}
AP <- MakeAgePrior(MaxAgeParent = c(12, 8), Plot=FALSE)
AP[as.character(seq(1,12,by=2)), c("MS", "FS")] <- 0
PlotAgePrior(AP)

# check if valid ageprior:
chk <- sequoia:::CheckAP(AP) 
```


## From age-difference distribution
If you know approximately the distributions of parental ages, and the age differences between siblings, you might be able to create a table similar to `tblA.R` and follow the subsequent steps described [here](#sec:tblAR) to generate an ageprior matrix. The most challenging may be to generate the age-difference distribution of 'all individuals' (column X), used for standardisation. 


## Specifications
The ageprior must abide by the following rules:

- a matrix or dataframe with values between 0 and 1000 inclusive. In the `MakeAgePrior()` output, the smallest possible non-zero value is 1/1000, to avoid excessive weight of the ageprior relative to the genetic likelihood;
- columns named M, P, FS, MS, and PS, any other columns are ignored;
- each relationship must be possible ($>0$) for at least one age difference;
- rownames are ignored and rows are presumed to be for age difference 0, 1, 2, ..., *except* when rownames include negative numbers (as for `AgePriorExtra`), then negative-numbered rows are deleted and the remainder are presumed to be ordered and for age difference 0, 1, 2, ...;
- any age differences not included (i.e. > `nrows` -1) are presumed impossible for all types of parents and siblings ($0$ in all 5 columns);
- the maximum age difference between siblings must be consistent with the maximum breeding tenure of a parent (minimum -- maximum age of reproduction).

To check whether your ageprior matrix satisfies these criteria, use it in any function that takes `AgePriors` (or `SeqList`) as input, or call the internal function `CheckAP()` directly.

# Estimate birth years
The estimated the distribution of age differences among relatives (i.e. ageprior matrix) can be used to estimate the birth year of individuals for which this is unknown. A requirement is that these focal individuals have one or more parents, offspring, or siblings with known birth year or BY.min + BY.max; the more relatives and the more exact their birth years, the better the birth year estimate of the focal individual. This can be done with function `CalcBYprobs()`, with any pedigree.                      

