# Default ageprior  {#sec:Default}
 -------------------------------

The default ageprior used during parentage assignment is flat and minimally informative: it only specifies that parent and offspring cannot have an age difference of 0 ($\alpha_{A,R}=0$ for $A=0$ and $R=M$ or $R=P$, black in Figure \@ref(fig:AP-default)) and that all other age / relationship combinations are allowed ($\alpha_{A,R}=1$, pale green). 

```{r AP-default, fig.cap="Default minimally informative ageprior", out.width = "70%"}
MakeAgePrior(LifeHistData = LH_HSg5)
```


The range of possible age differences is taken from `LifeHistData`: 

```{r AP-default-2}
table(LH_HSg5$BirthYear)
```

The maximum age of parents `MaxAgeParent` is initially set to the maximum age difference possible among individuals in `LifeHistData` (here 2005-2000 = 5 years) $+1$. This ensures that all pairs are age-wise considered as potential siblings [^a]; the age difference between siblings can never be larger than `MaxAgeParent` $-1$. 


[^a]: Up to version 2.0, initial `MaxAgeParent` was  set to `diff(BYrange)`, but sibling relationships were also allowed at this age difference, so this change (probably) won't affect parentage assignment or full pedigree reconstruction.  


