# AgePrior matrix explained
 ----------------------------
 
The ageprior matrix has 5 columns (more in sequoia versions before 2.0), corresponding to parent-offspring and sibling pairs (abbreviations in Table \@ref(tab:RelAbbr)). The number of rows equals at least the maximum age of parents (`MaxAgeParent`) plus one: the first row is for individuals born in the same year (age difference $A=0$), the second row for individuals born one year apart, etc (indicated by the rownames). 


```{r AP-grif1, out.width="70%"}
MakeAgePrior(Pedigree = Ped_griffin, Smooth=FALSE, Flatten=FALSE)  # Note: Ped_griffin includes a birth year column
```


A value of $0$ (black squares in the plot) indicates that an age difference / relationship combination is impossible; in the example here this is the case for among others mother-offspring pairs (`M`, first column) and age differences of 0 and 4. Any age differences not in this matrix (i.e. >4 years) are considered impossible ($0$) for all five relationships. Values greater than $1$ (green) indicate a relationship is more likely for that age difference than for a random pair, while values below $1$ (grey) indicate the relationship is less likely. 


```{r RelAbbr, echo=FALSE}
as.DF <- function(V, cn=c("name", "x")) {  # named vector --> data.frame
  setNames(data.frame(c1 = names(V), c2=V), cn)
}

Rels <- c(M = "Mother - offspring",
          P = "Father - offspring",
          FS = "Full siblings",   # From version XX
          MS = "Maternal siblings (full + half)",
          PS = "Paternal siblings (full + half)",
          MGM = "Maternal grandmother",
          PGF = "Paternal grandfather",
          MGF = "Maternal grandfather (+ paternal grandmother)",
          UA = "avuncular (niece/nephew -- aunt/uncle, full + half)",
          "(M/P)(M/P/F)A" = "Avuncular; Mother's/Father's Maternal/Paternal/Full sibling")

Rels <- cbind(as.DF(Rels, c("Column (R)", "Meaning")),
              "Version" = c(rep("all", 2), "from 1.0", "all", "all",
                            rep("up to 1.3 *", 3), "up to 1.3",
                            "from 2.0 *"))

knitr::kable(Rels, row.names=FALSE, booktabs=TRUE,
             caption = "AgePriors column names per sequoia version. *: From sequoia v2.0 only in AgePriorExtra")
``` 



## `AgePriorExtra`
The agepriors for grand-parental en avuncular pairs are calculated from the agepriors for parent-offspring and sibling relationships (details in section [Grandparents & Avuncular](#sec:GP-AU)). These extra columns are not calculated by `MakeAgePrior()`, but  by `sequoia()` just prior to parentage assignment and prior to full pedigree reconstruction (see pipeline in Figure \@ref(fig:pipeline)). 


```{r AP-grif2}
# sequoia() output for the griffin data is included in the package:
# see ?SeqOUT_griffin on how it was generated
round(SeqOUT_griffin$AgePriorExtra, 2)
PlotAgePrior(SeqOUT_griffin$AgePriorExtra)
```


The ageprior for avuncular pairs is complicated: aunts and uncles are often older than their nieces/nephews, but may also be younger, and the distribution is not necessarily symmetrical around zero. To accommodate this, age differences in `AgePriorExtra` are no longer absolute age differences, but do go negative. Parents and grandparents must always be older than their (grand-)offspring, and the distribution for siblings is always strictly symmetrical.  

